---
title: "Project: Default of Credit Card Clients"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  md_document:
  word_document: default
editor_options: 
  chunk_output_type: console
---

```{r}
# Importing credit card clients data from a CSV file hosted online.
# The 'read.csv()' function is used to read the data from the specified URL.
# 'header=T' indicates that the first row of the CSV file contains column names.
credit_default <- read.csv(file = "https://xiaorui.site/Data-Mining-R/lecture/data/credit_default.csv", header = TRUE)
colnames(credit_default)
table(credit_default$default.payment.next.month)
```

```{r}
library(dplyr)
credit_default <- rename(credit_default, default='default.payment.next.month')
table(credit_default$PAY_0)
```

## Constructing new variables

```{r}
# Create a new variable named 'BILL_AMT1_to_3' by summing up the bill amounts 
# for the first three months (BILL_AMT1, BILL_AMT2, and BILL_AMT3) for each observation
credit_default2 <- 
  mutate(credit_default, 
         BILL_AMT1_to_3 = BILL_AMT1 + BILL_AMT2 + BILL_AMT3)

summary(credit_default$BILL_AMT1_to_3)
```

## Creating sub-samples for specific needs

```{r}
# Create a subset 'credit_single' containing observations for clients who are single
credit_single <- 
  filter(credit_default, MARRIAGE == 2)

# Create a subset 'credit_married' containing observations for clients who are married
credit_married <- 
  filter(credit_default, MARRIAGE == 1)


# Generate summary statistics for the 'PAY_AMT1' variable for single clients
summary(credit_single$PAY_AMT1)

# Generate summary statistics for the 'PAY_AMT1' variable for married clients
summary(credit_married$PAY_AMT1)
```

# Exploratory Analyses & Visualization

### 1. Exploring the distribution of the dependent/independent variables

```{r}
summary(credit_default)

# Draw a pie chart to show the proportion of default clients
pie(table(credit_default$default))
```

> Question: Is this pie chart lacking appeal and clarity? How can we enhance its visual presentation? Our robot tutor is ready to assist you; just ask the right question!
> 
> "make this pie chart fancier: pie(table(credit_default$default))"

```{r}
# Create a pie chart with customized colors, labels, and a title
default_table <- table(credit_default$default)
colors <- c("#FF9999", "#66B2FF")  # Define custom colors for the slices
labels <- c("Not Default", "Default")  # Define custom labels for the slices
percentages <- round(default_table / sum(default_table) * 100, 1)  # Calculate percentages
pie(default_table, 
    col = colors, 
    labels = paste(labels, "\n (", percentages, "%)"),  # Include percentages in labels
    main = "Credit Default Status")
legend("topright", legend = labels, fill = colors, cex = 0.8)
```

```{r}
# Create a histogram to visualize the distribution of ages in the dataset
hist(credit_default$AGE, main = "Distribution of Age")

# Create a histogram to visualize the distribution of balance limits in the dataset
hist(credit_default$LIMIT_BAL, main = "Distribution of Balance Limit")
```

```{r}
# Generate a table to display the frequency of each education level in the dataset
table(credit_default$EDUCATION)

# Create a bar plot to visualize the frequency of each education level
barplot(table(credit_default$EDUCATION))

# Generate a table to display the frequency of each gender in the dataset
table(credit_default$SEX)
```

### 2. Exploring the association between default indicator and other numerical variables

```{r}
# Exploring the association between default indicator and other numerical variables
boxplot(credit_default$AGE ~ credit_default$default, horizontal = T)
boxplot(credit_default$BILL_AMT1 ~ credit_default$default, horizontal = T)
```

### 3. Exploring the association between default indicator and categorical variables

Exploring the association between a binary variable (such as a default indicator) and a categorical variable involves several analytical methods and visualization techniques. For example, frequency tables, bar charts, 

```{r}
# 3. Exploring the association between Salary and categorical variables
table(credit_default$default, credit_default$SEX, 
      dnn = c("Default or not","Sex: Gender (1 = male; 2 = female)")) 

# Create a contingency table of default status and marital status
myNewTable <- table(credit_default$default, credit_default$MARRIAGE) 

# Convert the contingency table into proportions
prop.table(myNewTable) 

# Create a stacked bar chart
barplot(myNewTable, 
        col=c('blue','red'),  # Define colors for the bars
        legend= paste("Default = ", rownames(myNewTable)),  
        # Add legend with default status labels
        xlab="Marital status",  # Label for x-axis
        ylab = "Count")  # Label for y-axis
```

```{r, eval=FALSE, echo=FALSE}
# Load the 'vcd' package for mosaic plot function
# install.packages("vcd")
library(vcd)

# Create a mosaic plot
mosaicplot(table(credit_default$MARRIAGE, credit_default$default), 
           main = "Mosaic Plot of Default vs. MARRIAGE",
           color = c("blue", "red"))  # Optional: Customize colors
```


# Building Models

## Preparing the training and testing datasets

```{r}
# Normalize the numerical variables 
var_used <- c("LIMIT_BAL", "AGE", 
              "PAY_0", "PAY_2", "PAY_3", 
              "PAY_4", "PAY_5", "PAY_6",
              "BILL_AMT1", "BILL_AMT2", "BILL_AMT3",
              "BILL_AMT4", "BILL_AMT5", "BILL_AMT6",
              "PAY_AMT1", "PAY_AMT2", "PAY_AMT3",
              "PAY_AMT4", "PAY_AMT5", "PAY_AMT6")

# Calculate the column-wise maximum ('maxs') and minimum ('mins') values for all numerical columns, excluding the first three columns.
maxs <- apply(credit_default[,var_used], 2, max)
mins <- apply(credit_default[,var_used], 2, min)

# Normalize the numerical data using min-max scaling and update 'credit_single_scaled'.
# - 'center = mins' ensures the data is centered to start at its minimum values.
# - 'scale = maxs - mins' scales the data to range from 0 to 1.
credit_default_scaled <- credit_default
credit_default_scaled[,var_used] <- 
  as.data.frame(scale(credit_default[,var_used], 
                      center = mins, 
                      scale = maxs - mins))

# Split the data into training and testing sets:
# - 'sample_index' contains randomly selected row indices for training, covering 70% of the entire dataset.
# - 'credit_train' contains the training subset of 'credit_single_scaled'.
# - 'credit_test' contains the remaining 30% as the testing subset.
# sample_index <- sample(nrow(credit_single_scaled), nrow(credit_single_scaled) * 0.70)
# credit_train <- credit_single_scaled[sample_index,]
# credit_test <- credit_single_scaled[-sample_index,]

# Split training and testing datasets to make sure static results
credit_train <- credit_default_scaled[1:8400,]
credit_test <- credit_default_scaled[8401:12000,]
table(credit_train$default)
table(credit_test$default)
```

## First Method: Logistic regression models

### A simple model based on your intuition: only select a few relevant predictors 

```{r}
# Build a model based on your intuition: only select a few relevant predictors 
DefaultModel1 <- 
  glm(default ~ LIMIT_BAL + SEX + MARRIAGE + AGE + PAY_0 + BILL_AMT1 + PAY_AMT1,
     data = credit_train)

summary(DefaultModel1, digits = 3)
```

### A simple model without imposing any constraints on predictor selection

```{r}
# Build a simple model without imposing any constraints on predictor selection
DefaultModelFull <- 
  glm(default ~ .,
     data = credit_train, 
     family = "binomial")

summary(DefaultModelFull, digits = 3)
```

### Advanced model development: Interaction variables

```{r}
# A potential model with interaction term
DefaultModel_inter <- 
  glm(default ~ . + AGE * MARRIAGE,
     data = credit_train, 
     family = "binomial")

summary(DefaultModel_inter, digits = 3)
```

### Advanced model development: nonlinear term

```{r}
# Fit a logistic regression model with quadratic age term
DefaultModel_quad <- 
  glm(default ~ . + I(AGE^2),  # Formula including quadratic term for age
     data = credit_train,  # Dataset
     family = "binomial")  # Binomial family for binary outcome
summary(DefaultModel_quad, digits = 3)  # Summary of model with 3 significant digits


# Fit a logistic regression model with quadratic PAY_AMT1 term
DefaultModel_quad2 <- 
  glm(default ~ . + I(PAY_AMT1^2),  # Formula including quadratic term for PAY_AMT1
     data = credit_train,  # Dataset
     family = "binomial")  # Binomial family for binary outcome
summary(DefaultModel_quad2, digits = 3)  # Summary of model with 3 significant digits
```

### Evaluating the logistic regression model 

1. In-sample AUC-ROC (or training set)

```{r}
# Predicted probability for training set
pred_train_logit <- predict(DefaultModel_quad2, newdata = credit_train, type="response")

library(ROCR)
# This line of code creates a prediction object 'pred' using the 'prediction()' 
pred_obj_train_logit <- prediction(pred_train_logit, credit_train$default)

# This line of code calculates True Positive Rate (TPR) and False Positive Rate
# (FPR) based on the prediction object 'pred_obj_train' using the 'performance()'
# function from the 'ROCR' package. 
# It stores the resulting performance object in the variable 'perf_train'.
perf_train_logit <- performance(pred_obj_train_logit, "tpr", "fpr")

# Draw the ROC curve
plot(perf_train_logit, colorize=TRUE)

# This line of code calculates the Area Under the ROC Curve (AUC) using the 'performance()' function with "auc" as the argument. 
# It extracts the AUC value from the resulting performance object and returns it as
# a numeric value.
# The 'unlist()' function is used to convert the AUC value from a list to a numeric vector.
unlist(slot(performance(pred_obj_train_logit, "auc"), "y.values"))
```

2. Out-of-sample AUC-ROC performance (testing set is more important)

```{r}
# Predicted probability for testing set
pred_test_logit <- predict(DefaultModel_quad2, newdata = credit_test, type="response")

library(ROCR)
# This line of code creates a prediction object 'pred' using the 'prediction()' 
pred_obj_test_logit <- prediction(pred_test_logit, credit_test$default)

perf_test_logit <- performance(pred_obj_test_logit, "tpr", "fpr")

# Draw the ROC curve
plot(perf_test_logit, colorize=TRUE)

# This line of code calculates the Area Under the ROC Curve (AUC) using the 'performance()' function with "auc" as the argument. 
unlist(slot(performance(pred_obj_test_logit, "auc"), "y.values"))
```


## Second Method: Neural Network

```{r}
library(neuralnet)
default_nn1 <- 
  neuralnet(formula = default ~ ., 
            data = credit_train, 
            hidden = c(3), 
            linear.output = FALSE)

plot(default_nn1)
```

### Evaluating the neural network model

We calculate the AUC-ROC for the training set. 

```{r}
# Predicted probability for training set
pred_train_nn1 <- predict(default_nn1, newdata = credit_train, type="response")

library(ROCR)
# This line of code creates a prediction object 'pred' using the 'prediction()' 
pred_obj_train_nn1 <- ROCR::prediction(pred_train_nn1, credit_train$default)

# This line of code calculates True Positive Rate (TPR) and False Positive Rate
# (FPR) based on the prediction object 'pred_obj_train' using the 'performance()'
# function from the 'ROCR' package. 
# It stores the resulting performance object in the variable 'perf_train'.
perf_train_nn1 <- performance(pred_obj_train_nn1, "tpr", "fpr")

plot(perf_train_nn1, colorize=TRUE)

# This line of code calculates the Area Under the ROC Curve (AUC) using the 'performance()' function with "auc" as the argument. 
# It extracts the AUC value from the resulting performance object and returns it as
# a numeric value.
# The 'unlist()' function is used to convert the AUC value from a list to a numeric vector.
unlist(slot(performance(pred_obj_train_nn1, "auc"), "y.values"))
```


2. Out-of-sample AUC-ROC performance (testing set is more important)

```{r}
# Predicted probability for testing set
pred_test_nn1 <- predict(default_nn1, newdata = credit_test, type="response")

library(ROCR)
# This line of code creates a prediction object 'pred' using the 'prediction()' 
pred_obj_test_nn1 <- ROCR::prediction(pred_test_nn1, credit_test$default)

perf_test_nn1 <- performance(pred_obj_test_nn1, "tpr", "fpr")

# Draw the ROC curve
plot(perf_test_nn1, colorize=TRUE)

# This line of code calculates the Area Under the ROC Curve (AUC) using the 'performance()' function with "auc" as the argument. 
unlist(slot(performance(pred_obj_test_nn1, "auc"), "y.values"))
```


# Conclusion

## Compare your two groups of methods

The logistic regression model `DefaultModel_quad2` performs less effectively out-of-sample compared to the neural network model `default_nn1`. Specifically, the logistic regression model achieves an AUC of 0.731 on the testing set, whereas the neural network model achieves a higher AUC of 0.758 on the same testing dataset.

## Business Recommendations

1. Monitor early signs of repayment delays
Recent repayment status was one of the strongest predictors of default. Financial institutions should implement early warning systems to identify customers who begin missing or delaying payments so that proactive support or payment plans can be offered.

2. Incorporate predictive models into credit risk systems
The neural network model achieved higher predictive performance than the logistic regression model, suggesting that machine learning approaches may better capture complex financial behavior patterns. These models could help lenders identify high-risk customers and improve credit decision-making.

3. Focus on behavioral financial indicators
Repayment history and billing behavior were stronger predictors of default than demographic variables. Credit risk assessments should prioritize behavioral financial data when evaluating customer risk.

## Limitations of your analysis

1. Limited model tuning
The models used relatively simple configurations without extensive hyperparameter tuning, which may limit their predictive performance.

2. Dataset scope
The dataset represents clients from a specific institution, so the findings may not fully generalize to other financial populations.

3. Missing external factors
Important variables such as income, employment status, or broader economic conditions were not included, which may affect the model’s ability to explain default risk fully.
