---
title: "Project: Default of Credit Card Clients"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  md_document:
  word_document: default
editor_options: 
  chunk_output_type: console
---

## Setup and data

Place `credit_default.csv` beside this notebook. Required packages are `dplyr`,
`ROCR`, `neuralnet`, `knitr`, and `rmarkdown`.

```{r setup}
knitr::opts_chunk$set(error = FALSE, fig.width = 8, fig.height = 5)
library(dplyr)
library(ROCR)
library(neuralnet)
credit_default <- read.csv("credit_default.csv")
credit_default <- rename(credit_default, default = default.payment.next.month)
stopifnot(!anyNA(credit_default), all(credit_default$default %in% 0:1))
# Explicit categories preserve nominal rather than numeric relationships.
credit_default$SEX <- factor(credit_default$SEX, levels = c(1, 2),
                             labels = c("Male", "Female"))
credit_default$EDUCATION <- factor(credit_default$EDUCATION, levels = 1:4,
                                  labels = c("Graduate", "University", "HighSchool", "Other"))
credit_default$MARRIAGE <- factor(credit_default$MARRIAGE, levels = 1:3,
                                 labels = c("Married", "Single", "Other"))
stopifnot(!anyNA(credit_default))
table(credit_default$default)
```

Repayment-status variables retain their original numeric codes. The supplied
notebook documents −1 as payment made duly and positive values as months of
payment delay, but does not establish the definitions of −2 and 0. These codes
are retained without assigning new meanings; numeric modeling assumes a common
ordered scale and should be compared with categorical encoding in future work.

## Constructing new variables

```{r}
# Create a new variable named 'BILL_AMT1_to_3' by summing up the bill amounts 
# for the first three months (BILL_AMT1, BILL_AMT2, and BILL_AMT3) for each observation
credit_default2 <- 
  mutate(credit_default, 
         BILL_AMT1_to_3 = BILL_AMT1 + BILL_AMT2 + BILL_AMT3)

summary(credit_default2$BILL_AMT1_to_3)
```

## Creating sub-samples for specific needs

```{r}
# Create a subset 'credit_single' containing observations for clients who are single
credit_single <- 
  filter(credit_default, MARRIAGE == "Single")

# Create a subset 'credit_married' containing observations for clients who are married
credit_married <- 
  filter(credit_default, MARRIAGE == "Married")


# Generate summary statistics for the 'PAY_AMT1' variable for single clients
summary(credit_single$PAY_AMT1)

# Generate summary statistics for the 'PAY_AMT1' variable for married clients
summary(credit_married$PAY_AMT1)
```

# Exploratory Analyses & Visualization

### 1. Exploring the distribution of the dependent/independent variables

```{r}
summary(credit_default)

# Draw a pie chart to show the proportion of default clients
pie(table(credit_default$default))
```

```{r}
# Create a pie chart with customized colors, labels, and a title
default_table <- table(credit_default$default)
colors <- c("#FF9999", "#66B2FF")  # Define custom colors for the slices
labels <- c("Not Default", "Default")  # Define custom labels for the slices
percentages <- round(default_table / sum(default_table) * 100, 1)  # Calculate percentages
pie(default_table, 
    col = colors, 
    labels = paste(labels, "\n (", percentages, "%)"),  # Include percentages in labels
    main = "Credit Default Status")
legend("topright", legend = labels, fill = colors, cex = 0.8)
```

```{r}
# Create a histogram to visualize the distribution of ages in the dataset
hist(credit_default$AGE, main = "Distribution of Age")

# Create a histogram to visualize the distribution of balance limits in the dataset
hist(credit_default$LIMIT_BAL, main = "Distribution of Balance Limit")
```

```{r}
# Generate a table to display the frequency of each education level in the dataset
table(credit_default$EDUCATION)

# Create a bar plot to visualize the frequency of each education level
barplot(table(credit_default$EDUCATION))

# Generate a table to display the frequency of each gender in the dataset
table(credit_default$SEX)
```

### 2. Exploring the association between default indicator and other numerical variables

```{r}
# Exploring the association between default indicator and other numerical variables
boxplot(credit_default$AGE ~ credit_default$default, horizontal = T)
boxplot(credit_default$BILL_AMT1 ~ credit_default$default, horizontal = T)
```

### 3. Exploring the association between default indicator and categorical variables

Counts describe sample composition; within-group proportions describe observed default rates. 

```{r}
# Compare observed default status across categories
table(credit_default$default, credit_default$SEX, 
      dnn = c("Default or not","Sex: Gender (1 = male; 2 = female)")) 

# Create a contingency table of default status and marital status
myNewTable <- table(credit_default$default, credit_default$MARRIAGE) 

# Convert the contingency table into proportions
prop.table(myNewTable, margin = 2) 

# Create a stacked bar chart
barplot(myNewTable, 
        col=c('blue','red'),  # Define colors for the bars
        legend= paste("Default = ", rownames(myNewTable)),  
        # Add legend with default status labels
        xlab="Marital status",  # Label for x-axis
        ylab = "Count")  # Label for y-axis
```

# Building Models

## Preparing the training and testing datasets

The full customer sample is split randomly within each outcome class using a
fixed seed. Scaling is fitted on training rows only. Test values may lie outside
[0, 1]; they are not clipped. The exploratory bill-total variable is not modeled.

```{r prepare}
set.seed(20260915)
by_class <- split(seq_len(nrow(credit_default)), credit_default$default)
sample_index <- sort(unlist(lapply(by_class, function(i) sample(i, round(length(i) * 0.70)))))
train_raw <- credit_default[sample_index, ]
test_raw <- credit_default[-sample_index, ]
stopifnot(length(intersect(sample_index, setdiff(seq_len(nrow(credit_default)), sample_index))) == 0)
var_used <- c("LIMIT_BAL", "AGE", "PAY_0", "PAY_2", "PAY_3", "PAY_4", "PAY_5", "PAY_6",
              paste0("BILL_AMT", 1:6), paste0("PAY_AMT", 1:6))
fit_scaler <- function(data) {
  mins <- vapply(data[var_used], min, numeric(1))
  ranges <- vapply(data[var_used], max, numeric(1)) - mins
  ranges[ranges == 0] <- 1
  list(mins = mins, ranges = ranges)
}
apply_scaler <- function(data, scaler) {
  data[var_used] <- as.data.frame(scale(data[var_used], center = scaler$mins, scale = scaler$ranges))
  data
}
scaler <- fit_scaler(train_raw)
credit_train <- apply_scaler(train_raw, scaler)
credit_test <- apply_scaler(test_raw, scaler)
rbind(Training = table(credit_train$default), Testing = table(credit_test$default))
```

## First Method: Logistic regression models

### A simple model based on your intuition: only select a few relevant predictors 

```{r}
# Build a model based on your intuition: only select a few relevant predictors 
DefaultModel1 <- 
  glm(default ~ LIMIT_BAL + SEX + MARRIAGE + AGE + PAY_0 + BILL_AMT1 + PAY_AMT1,
     data = credit_train, family = "binomial")

summary(DefaultModel1, digits = 3)
```

### A simple model without imposing any constraints on predictor selection

```{r}
# Build a simple model without imposing any constraints on predictor selection
DefaultModelFull <- 
  glm(default ~ .,
     data = credit_train, 
     family = "binomial")

summary(DefaultModelFull, digits = 3)
```

### Advanced model development: Interaction variables

```{r}
# A potential model with interaction term
DefaultModel_inter <- 
  glm(default ~ . + AGE * MARRIAGE,
     data = credit_train, 
     family = "binomial")

summary(DefaultModel_inter, digits = 3)
```

### Advanced model development: nonlinear term

```{r}
# Fit a logistic regression model with quadratic age term
DefaultModel_quad <- 
  glm(default ~ . + I(AGE^2),  # Formula including quadratic term for age
     data = credit_train,  # Dataset
     family = "binomial")  # Binomial family for binary outcome
summary(DefaultModel_quad, digits = 3)  # Summary of model with 3 significant digits


# Fit a logistic regression model with quadratic PAY_AMT1 term
DefaultModel_quad2 <- 
  glm(default ~ . + I(PAY_AMT1^2),  # Formula including quadratic term for PAY_AMT1
     data = credit_train,  # Dataset
     family = "binomial")  # Binomial family for binary outcome
summary(DefaultModel_quad2, digits = 3)  # Summary of model with 3 significant digits
```

### Evaluating the logistic regression model 

1. In-sample AUC-ROC (or training set)

```{r}
# Predicted probability for training set
pred_train_logit <- predict(DefaultModel_quad2, newdata = credit_train, type="response")

library(ROCR)
# This line of code creates a prediction object 'pred' using the 'ROCR::prediction()' 
pred_obj_train_logit <- ROCR::prediction(pred_train_logit, credit_train$default)

# This line of code calculates True Positive Rate (TPR) and False Positive Rate
# (FPR) based on the prediction object 'pred_obj_train' using the 'performance()'
# function from the 'ROCR' package. 
# It stores the resulting performance object in the variable 'perf_train'.
perf_train_logit <- performance(pred_obj_train_logit, "tpr", "fpr")

# Draw the ROC curve
plot(perf_train_logit, colorize=TRUE)

# This line of code calculates the Area Under the ROC Curve (AUC) using the 'performance()' function with "auc" as the argument. 
# It extracts the AUC value from the resulting performance object and returns it as
# a numeric value.
# The 'unlist()' function is used to convert the AUC value from a list to a numeric vector.
unlist(slot(performance(pred_obj_train_logit, "auc"), "y.values"))
```

2. Out-of-sample AUC-ROC performance (testing set is more important)

```{r}
# Predicted probability for testing set
pred_test_logit <- predict(DefaultModel_quad2, newdata = credit_test, type="response")

library(ROCR)
# This line of code creates a prediction object 'pred' using the 'ROCR::prediction()' 
pred_obj_test_logit <- ROCR::prediction(pred_test_logit, credit_test$default)

perf_test_logit <- performance(pred_obj_test_logit, "tpr", "fpr")

# Draw the ROC curve
plot(perf_test_logit, colorize=TRUE)

# This line of code calculates the Area Under the ROC Curve (AUC) using the 'performance()' function with "auc" as the argument. 
unlist(slot(performance(pred_obj_test_logit, "auc"), "y.values"))
```


## Second Method: Neural Network

Categorical predictors are represented by indicator columns using the same
training-derived design specification for both partitions. The network retains
one hidden layer with three neurons.

```{r neural-network}
nn_terms <- terms(~ ., data = credit_train[, setdiff(names(credit_train), "default")])
x_train <- model.matrix(nn_terms, credit_train)[, -1, drop = FALSE]
x_test <- model.matrix(nn_terms, credit_test)[, -1, drop = FALSE]
colnames(x_train) <- make.names(colnames(x_train), unique = TRUE)
colnames(x_test) <- make.names(colnames(x_test), unique = TRUE)
stopifnot(identical(colnames(x_train), colnames(x_test)))
nn_train <- data.frame(default = credit_train$default, x_train)
set.seed(20260916)
default_nn1 <- neuralnet(default ~ ., data = nn_train,
                         hidden = 3, linear.output = FALSE)
if (is.null(default_nn1$result.matrix)) stop("Neural network did not converge; no results reported.")
plot(default_nn1)
```

### Evaluating the neural network model

We calculate the AUC-ROC for the training set. 

```{r}
# Predicted probability for training set
pred_train_nn1 <- as.vector(predict(default_nn1, newdata = as.data.frame(x_train)))

library(ROCR)
# This line of code creates a prediction object 'pred' using the 'ROCR::prediction()' 
pred_obj_train_nn1 <- ROCR::prediction(pred_train_nn1, credit_train$default)

# This line of code calculates True Positive Rate (TPR) and False Positive Rate
# (FPR) based on the prediction object 'pred_obj_train' using the 'performance()'
# function from the 'ROCR' package. 
# It stores the resulting performance object in the variable 'perf_train'.
perf_train_nn1 <- performance(pred_obj_train_nn1, "tpr", "fpr")

plot(perf_train_nn1, colorize=TRUE)

# This line of code calculates the Area Under the ROC Curve (AUC) using the 'performance()' function with "auc" as the argument. 
# It extracts the AUC value from the resulting performance object and returns it as
# a numeric value.
# The 'unlist()' function is used to convert the AUC value from a list to a numeric vector.
unlist(slot(performance(pred_obj_train_nn1, "auc"), "y.values"))
```


2. Out-of-sample AUC-ROC performance (testing set is more important)

```{r}
# Predicted probability for testing set
pred_test_nn1 <- as.vector(predict(default_nn1, newdata = as.data.frame(x_test)))

library(ROCR)
# This line of code creates a prediction object 'pred' using the 'ROCR::prediction()' 
pred_obj_test_nn1 <- ROCR::prediction(pred_test_nn1, credit_test$default)

perf_test_nn1 <- performance(pred_obj_test_nn1, "tpr", "fpr")

# Draw the ROC curve
plot(perf_test_nn1, colorize=TRUE)

# This line of code calculates the Area Under the ROC Curve (AUC) using the 'performance()' function with "auc" as the argument. 
unlist(slot(performance(pred_obj_test_nn1, "auc"), "y.values"))
```


# Model evaluation summary

The logistic specification was retained from the original project rather than
selected using test performance. A threshold of 0.5 is illustrative and was not
tuned on the test set. Brier score measures squared probability error (lower is
better); the calibration table compares average predictions with observed rates.

```{r metrics}
auc_value <- function(p, y) as.numeric(performance(ROCR::prediction(p, y), "auc")@y.values[[1]])
metrics <- function(p, y) {
  stopifnot(length(p) == length(y), all(is.finite(p)), all(p >= 0 & p <= 1))
  labels <- as.integer(p >= 0.5)
  tp <- sum(labels == 1 & y == 1); fp <- sum(labels == 1 & y == 0)
  fn <- sum(labels == 0 & y == 1)
  c(AUC = auc_value(p, y), Accuracy = mean(labels == y),
    Recall = tp / (tp + fn), Precision = if (tp + fp == 0) NA_real_ else tp / (tp + fp),
    Brier = mean((p - y)^2))
}
results <- rbind(
  Logistic_training = metrics(pred_train_logit, credit_train$default),
  Logistic_testing = metrics(pred_test_logit, credit_test$default),
  Neural_network_training = metrics(pred_train_nn1, credit_train$default),
  Neural_network_testing = metrics(pred_test_nn1, credit_test$default))
knitr::kable(results, digits = 4)
calibration <- function(p, y) {
  bins <- cut(p, breaks = seq(0, 1, 0.1), include.lowest = TRUE)
  aggregate(data.frame(Predicted = p, Observed = y), list(Probability_bin = bins), mean)
}
calibration(pred_test_logit, credit_test$default)
calibration(pred_test_nn1, credit_test$default)
```

## Logistic stability within the training data

Five stratified folds assess the retained logistic specification without using
the held-out test set. Scaling is fitted separately within each fold. This is
not a cross-validation comparison of the two model families; the neural network
is evaluated using a single seeded holdout run.

```{r cross-validation}
set.seed(20260917)
fold <- integer(nrow(train_raw))
for (value in 0:1) {
  idx <- which(train_raw$default == value)
  fold[idx] <- sample(rep(1:5, length.out = length(idx)))
}
cv_auc <- vapply(1:5, function(k) {
  fold_train <- train_raw[fold != k, ]; fold_valid <- train_raw[fold == k, ]
  fold_scaler <- fit_scaler(fold_train)
  fold_train <- apply_scaler(fold_train, fold_scaler)
  fold_valid <- apply_scaler(fold_valid, fold_scaler)
  fit <- glm(default ~ . + I(PAY_AMT1^2), data = fold_train, family = binomial())
  auc_value(predict(fit, fold_valid, type = "response"), fold_valid$default)
}, numeric(1))
data.frame(Fold = 1:5, AUC = cv_auc)
```

# Conclusion

On the held-out test set, logistic regression achieved AUC
`r sprintf("%.4f", results["Logistic_testing", "AUC"])`, and the neural network
achieved AUC `r sprintf("%.4f", results["Neural_network_testing", "AUC"])`.
These values are generated from the current run. Higher AUC indicates better
ranking of defaulting versus non-defaulting customers on this split; it does
not demonstrate why one model performs better or establish consistent superiority.

The mean five-fold training-validation AUC for the logistic specification was
`r sprintf("%.4f", mean(cv_auc))` (fold standard deviation
`r sprintf("%.4f", sd(cv_auc))`).

## Business interpretation and limitations

The models provide a starting point for studying default-risk scoring. This
analysis does not establish which predictors are strongest or demonstrate that
behavioral variables outperform demographics. Operational use would require
validation on representative current customers, threshold selection based on
error costs, and further calibration assessment.

The network uses one initialization and a small architecture without systematic
tuning. Its holdout result is not a repeated cross-validation estimate. The
historical sample may not generalize to other institutions or periods, and
income, employment, and broader economic conditions are not included.
Repayment codes are modeled numerically, an assumption worth testing separately.

## Reproducibility

```{r save-results}
dir.create("results", showWarnings = FALSE)
write.csv(data.frame(Model = rownames(results), results, row.names = NULL),
          "results/model_metrics.csv", row.names = FALSE)
write.csv(data.frame(Fold = 1:5, AUC = cv_auc), "results/logistic_cv.csv", row.names = FALSE)
write.csv(data.frame(Row = seq_len(nrow(credit_default)),
                     Partition = ifelse(seq_len(nrow(credit_default)) %in% sample_index, "Training", "Testing")),
          "results/data_split.csv", row.names = FALSE)
writeLines(capture.output(sessionInfo()), "results/session_info.txt")
sessionInfo()
```
